---
title: "Lesson 17"
author: "Melinda K. Higgins, PhD."
date: "October 24, 2017"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
knitr::opts_chunk$set(message = FALSE)
knitr::opts_chunk$set(warning = FALSE)
```

## Lesson 17 - Analysis of Covariance (ANCOVA)

In this lesson, we're working with the `HELP` dataset. We will be using these variables:

* `sexrisk` - dependent variable (DV)
* `mcs` - independent variable (IV)
* `pss_fr` - continuous covariate (CV)
* `female` - discrete/categorical covariate (CV)

Load the initial packages `tidyverse` and `haven` and load the HELP dataset.

```{r}
# load packages
library(tidyverse)
library(haven)

# load dataset
help1 <- haven::read_spss("helpmkh.sav")
```

Select the variables we want for this lesson: 

* DV: `sexrisk`
* IV: `mcs`
* CV: `pss_fr` or `female`

Let's also recode `female` to `male` by flipping the 0,1 and let's mean center `mcs` and `pss_fr` since we'll also look at the interaction between these variables plus the interactions between `mcs` and gender (`female` or `male`).

```{r}
help2 <- help1 %>%
  select(sexrisk, mcs, pss_fr, female) %>%
  mutate(mcsC = mcs - mean(mcs),
         pss_frC = pss_fr - mean(pss_fr),
         male = as.numeric((female==0)),
         mcsC_x_pss_frC = mcsC * pss_frC,
         mcsC_x_female = mcsC * female,
         mcsC_x_male = mcsC * male)
```

## Regression Approach 

Using a regression approach - look at `pss_fr` as a covariate for the relationship between `sexrisk` and `mcs` adjusting for `pss_fr`. Note the use of `*` which computes the interaction for you.

### regression summary

```{r}
m1 <- lm(sexrisk ~ mcsC * pss_frC, data=help2)
summary(m1)
```

### `olsrr` regression summary

```{r}
library(olsrr)
ols_regress(m1)
```

## Using an ANOVA approach

Let's run this again using the `aov()` function. Look at usual `summary()` function and the `ols_regress()` function from the `olsrr` package.

```{r}
m1aov <- aov(sexrisk ~ mcsC * pss_frC, data=help2)
summary(m1aov)
ols_regress(m1aov)
```

## Effect Plots

Let's also make an effectplot showing the interaction or lack thereof between `mcs` and `pss_fr` for `sexrisk`. For this we will use the `effects` package by John Fox.

```{r}
library(effects)
plot(effects::effect("mcsC:pss_frC", m1, 
            xlevels=list()),
     multiline=TRUE, ylab="Sex Risk", rug=FALSE)
```

### adjust the y-axis to more reasonable range

```{r}
plot(effects::effect("mcsC:pss_frC", m1, 
            xlevels=list()),
     multiline=TRUE, ylab="Sex Risk", rug=FALSE,
     ylim=c(0,8))
```

## Consider gender as a covariate

Now, let's look at gender as a covariate. First, run for `female` and then run for `male`. Knowing which category is coded as 0 and which is coded as 1 changes the interpretation of the intercept but also changes the interpretation of the main effect of `mcs` which is _**conditional upon**_ gender.

### Model for `female` where male=0 and female=1

```{r}
m2f <- lm(sexrisk ~ mcsC * female, data=help2)
summary(m2f)
ols_regress(m2f)
```

### Model for `male` where male=1 and female=0

```{r}

m2m <- lm(sexrisk ~ mcsC * male, data=help2)
summary(m2m)
ols_regress(m2m)
```

## Run gender as a covariate using the `aov()` function

### ANCOVA using `female` variable

```{r}
m2f.aov <- aov(sexrisk ~ mcsC * female, data=help2)
summary(m2f.aov)
ols_regress(m2f.aov)
```

### Check type III SS using Anova() from car package

```{r}
library(car)
car::Anova(m2f.aov, type=3)
```

### ANCOVA using `male` variable

```{r}
m2m.aov <- aov(sexrisk ~ mcsC * male, data=help2)
car::Anova(m2m.aov, type=3)
```

## Other packages for plotting interaction effects

Let's look at a couple more packages for plotting interaction effects:

* `HH` package
* `ggplot2` package
* `sjPlot` package

### Using the `HH` package to run `ANCOVA` and associated plot

Note: to use the `ancova()` function in the `HH` package, the categorical variable `female` must be of factor class. So, I created a new variable `femaleF` which is now a `factor` class.

```{r}
library(HH)
class(help2$female)
help2$femaleF <- as.factor(help2$female)
class(help2$femaleF)

hhaov <- HH::ancova(sexrisk ~ mcsC * femaleF, data=help2)
# see summary and associated effect plot
hhaov
```

### Using the `ggplot2` package to visualize interaction effects

Note: This plot can also be done using `qplot()` from the `ggplot2` package with an overlay of a smooth regression fit line. 

Note: `ggplot2` should already be loaded
when we loaded the `tidyverse` package above

```{r}

qplot(x=mcsC, y=sexrisk, facets=~female, data=help2) +
  geom_smooth(method="lm")
```

### Using the `sjPlot` package to make interaction plots

Learn more about the `sjPlot` pacakge and specifically the `sjp.int()` function at [http://www.strengejacke.de/sjPlot/sjp.int/](http://www.strengejacke.de/sjPlot/sjp.int/)

**NOTICE** `sjPlot` is using the labels from the "codebook" applied and in use when we read the dataset in using the `tidyverse` compatible `haven` package.

```{r}
library(sjPlot)
m3f <- lm(sexrisk ~ female * mcsC, data=help2)
sjPlot::sjp.int(m3f, type="eff",
        show.ci=TRUE,
        facet.grid=TRUE)
```

